\section*{\centering Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

This dissertation represents the culmination of my work at Forschungszentrum Jülich, Bergische Universität Wuppertal, and Colorado State University in the period from 2015 until 2020.

I like to emphasize my gratitude to my supervisor Prof.\ Dr.\ Lukas Arnold, as well as Prof.\ Dr.\ Armin Seyfried as the head of my institute. Both have made my PhD project a success, and I am grateful for their continuous support and encouragement throughout the project. %In particular, I would like to thank both of them for making it possible for me to work abroad allowing me to make huge progress in the project that was mandatory for its completion.
Further, I wish to acknowledge Bundesministerium für Bildung und Forschung (BMBF) for financially supporting this project within the framework of ORPHEUS (funding code 13N13266).

The investigations carried out in this dissertation were performed with the \dealii{} library, and I cannot thank the whole \dealii{} community enough for all their support. In particular, I would like to express my gratitude towards Prof.\ Dr.\ Wolfgang Bangerth for hosting me during a seven month research stay at the Colorado State University and for reviewing this thesis, as well as Peter Munch for his interest in both my work and its expansion and for proofreading. I would like to also thank Dr.\ Denis Davydov and Dr.\ Martin Kronbichler for their collaboration, as well as Graham Harper and Sean McGovern for lots of \dealii{} related discussions.

Furthermore, I wish to thank all the people I had the pleasure of meeting and working with during my time at the Forschungszentrum Jülich and the Bergische Universität Wuppertal. Thanks to Dr.\ Anne Kathrin Küsters and My Linh Würzburger for helpful discussions about numerics and programming and for proofreading, as well as Dr.\ Alexander Belt for his continuous motivation. In addition, I would also like to mention Dr.\ Erik Andresen, Anna Braun, Tristan Hehnen, Susanne Hendel, Patrick Lauer, Jette Schumann, Alec Singh, Corinna Trettin, and Ashish Vinayak, who became close friends.

Last, I am grateful to all friends and family for their continuous support on my journey. Be it through hiking adventures, joint sports activities such as running, cycling and swimming, riding roller coasters, attending concerts or simply by spending time together in any other possible way. It is impossible to mention all the people I have in mind here, and I hope that you, as a reader of these lines, feel addressed and associate pleasant memories in the same way as I do. I am grateful to have spent these moments with you and look forward to many more to come. Thank you.

I dedicate this work to my parents Elisabeth and Jürgen Fehling who supported me in my decision to pursue a PhD, always stood behind me and encouraged me to persevere.