\section*{\centering Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

The investigations carried out in this dissertation were performed with the \dealii{} library, and I cannot thank the whole \dealii{} community enough for all their support. In particular, I would like to express my gratitude towards Prof.\ Dr.\ Wolfgang Bangerth for hosting me during a seven month research stay at the Colorado State University and for reviewing this thesis, as well as Peter Munch for his interest in both my work and its expansion and for proofreading. I would like to also thank Dr.\ Denis Davydov and Dr.\ Martin Kronbichler for their collaboration, as well as Graham Harper and Sean McGovern for lots of \dealii{} related discussions.

I like to emphasize my gratitude to Prof.\ Dr.\ Armin Seyfried as the head of my home institute and my supervisor Prof.\ Dr.\ Lukas Arnold for making my PhD project possible. I am grateful for their support and encouragement, especially in spending about half a year abroad which allowed me to make huge progress in the project that was mandatory for its completion.

Furthermore, I wish to thank all the people I had the pleasure of meeting and working with during my time at the Forschungszentrum J체lich and the Bergische Universit채t Wuppertal. Thanks to Dr.\ Anne Kathrin K체sters and My Linh W체rzburger for helpful discussions about numerics and programming and for proofreading, as well as Dr.\ Alexander Belt for his continuous motivation. In addition, I would also like to mention Dr.\ Erik Andresen, Anna Braun, Patrick Lauer, Jette Schumann, Corinna Trettin, and Ashish Vinayak, who became close friends.

Last, I am grateful to all friends and family for their continuous support on my journey. Be it through hiking adventures, joint sports activities such as running, cycling and swimming, riding roller coasters, attending concerts or simply by spending time together in any other possible way. It is impossible to mention all the people I have in mind here, and I hope that you, as a reader of these lines, feel addressed and associate pleasant memories in the same way as I do. I am grateful to have spent these moments with you and look forward to many more to come. Thank you.
