\chapter{Dynamic parallel \hp-adaptive \glsfmtlongpl{fem}}
\label{ch:dynamic}
\glsresetall

Up to this point, we have only dealt with static \hp-adaptive meshes, in which the grid resolution and the distribution of finite elements is prescribed from the beginning. % Those can be used for multi-physics applications, in which different characteristics can be assigned to certain parts of the domain using dedicated finite elements.
However, the key feature of \hp-adaptive methods is the dynamic distribution of these attributes. The goal is to reduce the overall error of the approximation on the basis of the complexity of the current state of the solution. With dynamic methods, meshes can be tailored to a specific problem iteratively with so called SOLVE-ESTIMATE-MARK-REFINE cycles: After solving the problem on a coarse mesh, it will be adapted based on an error estimate several times, as we will demonstrate in Ch.~\ref{ch:results}. Also, time dependent problems benefit from dynamic adaptation, in which mesh attributes will be adjusted on the basis of the solution that evolves in time, for example, for heat transfer \textcite{dealiistep-26} and fluid flow problems \textcite{dealiistep-31}.

In practice, we distinguish between two ways of imposing these \hp-adaptive methods: either manually or automatically. The combination of both \h- and \p-adaptation poses new challenges. We will present these issues in this chapter along with algorithms to automatically determine which regions to adapt and which type of adaptation to impose.

First, considerations regarding the implementation of \hp-adaptivity will be presented. We then propose methods to automatically decide which cells to adapt and how by providing corresponding adaptation and decision strategies. The presented strategies are intended for general purposes and are based on error estimation, error prediction and smoothness estimation. It is conceivable to tailor strategies around particular observables of individual problems, which is not subject of this dissertation. Furthermore, dynamic changes on the mesh require data exchange between processes, which we describe in the end.

New features for \hp-adaptation as well as a generalized interface have been introduced in the \dealii{} library \textcite{dealii920} during working on this dissertation. A development log on all features introduced in the context of dynamic \hp-adaptive \glspl{fem} can be found here \cite{dealiiissue7515}.



\input{chapters/dynamic/strategy}
\input{chapters/dynamic/decision}
\input{chapters/dynamic/transfer}